\vspace{-2.0cm}Stuff

### 2014-05-29 report/draft3.tex
Updated draft after meeting with Claudio.

\usepackage{fullpage}
\usepackage{floatrow}
\usepackage{fourier}
\usepackage{parskip}% http://ctan.org/pkg/parskip
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{textcomp}
%\usepackage[round,authoryear]{natbib}
\usepackage[utf8]{inputenc}
\usepackage[backend=biber,citestyle=authoryear,bibstyle=authoryear,maxbibnames=10]{biblatex}
\addbibresource{VCPbibliography.bib}

\begin{document}
Ignore this page, it's so I can have a reference for the different citation formats and will be removed before turning in the paper.

cite \cite{buckland2001} \cite{komodo2013} \\
cite* \cite*{buckland2001} \cite*{komodo2013}\\
parencite \parencite{buckland2001} \parencite{komodo2013}\\
parencite* \parencite*{buckland2001} \parencite*{komodo2013}\\
textcite \textcite{buckland2001} \textcite{komodo2013}\\
smartcite - does the super script/foot note thing.
\newpage
\begingroup  
  \centering
  \LARGE Master's Report - Variable Circular Plots: Station Placement and the Independence Assumption\\[1em]
  \large Matt Edwards\\
  26 May 2014\par
\endgroup

\section{Introduction}
Estimating population density of plants or animals is a persistent issue in the ecological sciences. Distance sampling, in the form of line and point transects (alternatively, Variable Circular Plots (VCP)), is a common estimation method. They are useful when individual objects are too small to be seen from the air or may be masked by the canopy (birds, small animals, ground plants.)

Distance sampling has been around since the early 1900's. In the 1970's, more rigorous scientific methods were applied, including papers by Ramsey and Scott \parencite*{ramsey1979,ramsey1981} which augmented work done by Emlen in 1971. \cite{burnham1980} is a monograph on the theory and application of line transect sampling. Non-parametric methods have been examined by \textcite{quang1993,mack1998}, among others. Distance sampling continues to be used and modified, combined with mark-recapture methods \parencite{laake2011}, extended for use with underwater acoustics for estimates of krill populations \parencite{krill2011}, and a study by \textcite{camera2011} combines distance sampling with camera traps to estimate populations.

\textcite{buckland2001} has a more in-depth discussion of the history of distance sampling.

@@

\textcite[6]{buckland2001} draws a distinction between point transects, where only the area immediately surrounding the point is surveyed (area of estimated perfect detectability), and Variable Circular Plots, where you census the area $\pi w^2$, where $w$ is the furthest distance at which you can observe the object of interest. However, in \textcite{buckland2006, quang1993} the terms are used interchangeably. They will be used interchangeably here also, with the understanding this paper addresses the second type, where an area of $pi*w^2$ is censused.

Objects closer to the observer are typically easier to spot than objects further away. Considering the probability of detecting an object, we can imagine that there is an area of (nearly) perfect detectability near the observer, that drops off the further away from the observer the object is. This detectability function is labeled $g(r)$, with $r$ being the observation distance. If no other covariates are tracked, we end up with a count of objects observed, and the distances from the line or point at which they were observed.

@@

An estimate of population density is a count of objects observed, divided by the area in which they were observed, times the probability of observing the object at that distance.

$\hat{D}=E[n]/(Area)*P(observing object|distance r)$

@@

\section{Micronesian Bird Data}
\begin{figure}
	\caption{Aggregated Collared Kingfisher Observation Distances, Rota \& Tinian, 20 m bins: \cite{micronesian}}
	\includegraphics[width=\textwidth]{../images/histogram_dist_20m.pdf}
	\label{fig:82dist}
\end{figure}
The Micronesian Forest Bird Survey \cite{micronesian} is a work using Variable Circular Plots (VCPs) placed using a random-systematic transect design to monitor the density of several bird species in the Micronesian islands.

@@

First, there are very low counts of birds immediately around the station, a peak at around 100 m, a fairly level section from 100--200 m, and a drop off after 200 m. This would seem to indicate that the presence of the observers caused either movement away from the station (the crest around 100m) or alternatively a suppression of vocalizations and/or movement directly around the station.

@@

One of the assumptions of distance sampling is that the objects being studied are distributed homogeneously with some uniform density, $D$, throughout the space \textcite{ramsey1981}.  (If the data is clustered we have other problems, not addressed here.) If the uniform distribution is the case, then it would make sense that it shouldn't matter if the observations overlap, as long as the act of observing at one station does not affect the observations at the neighboring station. This echoes Buckland’s statement a few paragraphs previous.

@@

\subsection{Bias \& Consistency}
\textcite{buckland2001,ramsey1979} Buckland et al. (2001) and Ramsey \& Scott (1970) both stated that if our detectability estimate $g(z)=$1 out to $r$, then $\hat{D}$ is unbiased.

\textcite{barry2001} Barry and Walsh (2001) compared design-based and  model-based estimates of density and variance. (Their focus was line transects, but they state it is analogous to point transects and the conclusions should carry across.) They concluded that model-based estimates were unbiased if and only if the assumptions of objects being independent and uniformly located held. They noted that in practice, this is probably not a feasible assumption. For design-based inference, they caution about both the density and variance estimates.

\textcite{roeder1987} Roeder, Dennis, \& Garton (1987) use simulations to demonstrate that no method is completely applicable in 100\% of situations, and will go sideways at some point, depending on the specific conditions.

The overall conclusion is that, in theory, density estimates using VCPs are unbiased and (reasonably) consistent, but that the assumptions necessary for the theory to hold are untenable. The appropriate estimation of $g(r)$ will vary depending on the specific situation, and it is up to the scientist and statistician to understand the data and choose the best approach.

\section{Simulation}
As seen in the previous section, the issue of independence of the point transects does not have a consensus in the literature. If independence is required for proper estimation, then the random-systematic placement (hereafter, ``transect'') layout may require some adjustment of the estimate to bring it in line with survey layouts where the observations areas do not overlap.

Through simulation, we can explore what happens under three different VCP placement schemes, detection densities, and object movement, and compare the resulting density estimates to their known values to evaluate the performance of the estimators.

\subsection{VCP Arrangement}
\label{sec:layout}
To examine the effect of VCP overlap, we will look at three options for VCP arrangement:
\begin{itemize}
	\item Structured: VCPs placed so that observation distances do not overlap
	\item Random: VCPs placed completely at random
	\item Transect: Two transects of 18 stations each, placed according to \textcite{micronesian} design: 150 m between stations, 2 km distance between parallel transects.
\end{itemize}
\begin{figure}
	\centering
	\caption{VCP Layout options. On graph, 1 unit = 1 km. Circles represent 200 m mark.}
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{../images/layout_structured.pdf}
		\caption{Structured Layout}
		\label{fig:structured}
	\end{subfigure}
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{../images/layout_random.pdf}
		\caption{Totally Random Layout}
		\label{fig:random}
	\end{subfigure}

	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{../images/layout_rand-sys-4.pdf}
		\caption{A Transect Layout}
		\label{fig:transect1}
	\end{subfigure}
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{../images/layout_rand-sys-6.pdf}
		\caption{Another Transect Layout}
		\label{fig:transect2}
	\end{subfigure}
	\label{fig:layouts}
\end{figure}
Figures~\ref{fig:structured} and \ref{fig:random} illustrate the first two arrangements. Density in these images is based on the 20 birds per km$^2$ of the Palie region on Rota.

@@

\subsection{$g(r)$ Options}
The first simulation compares the three layouts described in section~\ref{sec:layout} with two different options for $g(r)$, a half-normal detection probability function, and a detection probability estimated from the empirical observation distances.
\subsubsection{Half Normal}
For a maximum possible observation distance of 0.500 (500 m), the parameter for the half-normal detection function is $\theta = 8.77$, with a scale parameter $\delta = 0.179$.

@@

For a half-normal with parameter $\theta=8.7732$, $f(0)=5.5852$ so to scale the value to 1:

$\delta_{HN} = 1/5.5852 = 0.1790$

@@

\section{Results}
\subsection{Half-normal vs. Empirical $g(r)$}
\begin{figure}

@@

\texttt{TODO: Try to xplain the .50 jump in the estimate for transect/compounded. No theory or story atm.}

@@

\section{The Problem with Circles}
To test the effectiveness of my movement algorithm, I compared the detection distances from the data where the objects moved, to the detection distances from the same data, assuming they stayed still. The movement data, (left panel, Figure~\ref{fig:moveTest}) is squashed more towards 0.1 Km than the still data (right panel). But there's a problem. The still data doesn't show the nice shoulder at 0 that the literature instructs us to expect.

@@

The problem is because circles are different than rectangles.

With a line transect, as we move away from the line, the distance ``surveyed" increases linearly. With a point transect, as we move away from the point, the distance surveyed increases exponentially. Figure~\ref{fig:perfectArea} shows the relationship for a rectangle and circle with the same area.

Figure~\ref{fig:perfectCount} shows the difference in the expected number of observed objects at distance $r$ given a line transect vs. a VCP. For the VCP, Area was calculated as $2*pi*r$. For the line transect, Area was calculated as $(2r)*L$, with $L=0.785$ chosen so that the area of the circle and rectangle were equal when $r=0.5$.

@@

My conclusion from this is that for a VCP survey, even if everything is perfect, you will not see the nice shouldered layout of your empirical detection distances. \textcite{ramsey1981} speaks about this somewhat, when comparing the ``CUM-D'' curve, or the plot of the cumulative number of observations at distance $r$ for both line and point transects. However, almost all the illustrations on how your collected data should look, shows data that echoes the detection probability curve, with the half-bell-shape with a nice shoulder. Since this shape is impossible, I think some better examples are in order.

Better mathematicians than I have worked on distance sampling, and this is why in analyzing point transect data we take the derivative of $g(r)$, to adjust for this difference in how area is calculated.

@@

\begin{figure}
	\centering
	\caption{Rectangles (Line Transect, solid) vs. Circles (Point Transect, dashed)}
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{../images/rect-circ-area.pdf}
		\caption{Area of rectangle vs circle at same distance from point/line}
		\label{fig:perfectArea}
	\end{subfigure}
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{../images/rect-circ-detection.pdf}
		\caption{Estimated detection numbers for rectangle (solid) vs circle (dashed)}
		\label{fig:perfectCount}
	\end{subfigure}
\end{figure}

\section{Conclusions}
If all else is held equal, VCP layout does not seem to play a large role in density estimate using the kernel method as described by \textcite{quang1993}. Violations of the expected detection probability curve, or by movement of the objects away from the observer play more of a role in biasing the resulting estimate.

\section{Bibliography}

>>>>

\usepackage{fullpage}
\usepackage{floatrow}
\usepackage{fourier}
\usepackage{parskip}% http://ctan.org/pkg/parskip
\usepackage{graphicx}
\usepackage[labelfont={bf},margin=1cm]{caption}
\usepackage{subcaption}
\usepackage{textcomp}
%\usepackage[round,authoryear]{natbib}
\usepackage[utf8]{inputenc}
\usepackage[backend=biber,citestyle=authoryear,bibstyle=authoryear,maxbibnames=10]{biblatex}
\addbibresource{VCPbibliography.bib}

\AtEveryBibitem{ %
	\clearfield{url} %
	\clearfield{ISSN} %
	\clearfield{ISBN} %
}

%\usepackage{hyperref}

\begin{document}
%Ignore this page, it's so I can have a reference for the different citation formats and will be removed before turning in the paper.
%
%
%cite \cite{buckland2001} \cite{komodo2013} \\
%cite* \cite*{buckland2001} \cite*{komodo2013}\\
%parencite \parencite{buckland2001} \parencite{komodo2013}\\
%parencite* \parencite*{buckland2001} \parencite*{komodo2013}\\
%textcite \textcite{buckland2001} \textcite{komodo2013}\\
%smartcite - does the super script/foot note thing.
%\newpage
\begingroup  
  \centering
  \LARGE Masters Report - Variable Circular Plots: Station Placement and the Independence Assumption\\[1em]
  \large Matt Edwards\\
  26 May 2014\par
\endgroup

\section{Introduction}
Estimating population density of plants or animals is a persistent issue in the ecological sciences. Distance sampling, in the form of line and point transects (alternatively, Variable Circular Plots (VCP)), is a common estimation method. They are useful when individual objects are too small to be seen from the air or may be masked by the canopy (birds, small animals, ground plants, etc.).

Distance sampling has been around since the early 1900s. In the 1970s, more rigorous scientific methods were applied, including papers by Ramsey and Scott \parencite*{ramsey1979,ramsey1981} which augmented work done by Emlen in 1971. \cite{burnham1980} is a monograph on the theory and application of line transect sampling. Non-parametric methods have been examined by \textcite{quang1993,mack1998}, among others. Distance sampling continues to be used and modified, combined with mark-recapture methods \parencite{laake2011}, extended for use with underwater acoustics for estimates of krill populations \parencite{krill2011}, and a study by \textcite{camera2011} combines distance sampling with camera traps to estimate populations. For a more in-depth discussion of the history of distance sampling see \textcite{buckland2001}.

@@

\textcite{buckland2001} draws a distinction between point transects, where only the area immediately surrounding the point is surveyed (area of estimated perfect detectability), and Variable Circular Plots, where a census is taken of the area $\pi w^2$, where $w$ is the furthest distance at which there is a possibility of observing the object of interest. However, in \textcite{buckland2006, quang1993} the terms are used interchangeably. They will be used interchangeably here also, with the understanding this paper addresses the second type, where an area of $\pi w^2$ is censused.

Objects closer to the observer are typically easier to spot than objects further away. Considering the probability of detecting an object, we can imagine that there is an area of (nearly) perfect detectability near the observer, that drops off the further away from the observer the object is. This detectability function, defined as the probability of detecting an object, given that it is at distance $r$ is labeled $g(r)$, with $r$ being the observation distance. If no other covariates are tracked, we end up with a count of objects observed, and the distances from the line or point at which they were observed.

@@

An estimate of population density is $n$, the count of objects observed, divided by the area in which they were observed, times the probability of observing the object at that distance. \cite{buckland2001} %eq 2.12, pg 39

$\hat{D}=n/(Area)*P(observing object|distance r)$

@@

\section{Micronesian Bird Data}
\begin{figure}
	\caption{Aggregated Collared Kingfisher Observation Distances, Rota \& Tinian, 20 m bins: \cite{micronesian}}
	\includegraphics[width=\textwidth]{../images/histogram_dist_20m.pdf}
	\label{fig:82dist}
\end{figure}
The ``Micronesian Forest Bird Survey'' by \textcite{micronesian} is a work using Variable Circular Plots (VCPs) placed using a random-systematic transect design to monitor the density of several bird species in the Micronesian islands.

@@

First, there are very low counts of birds immediately around the station, a peak at around 100 m, a fairly level section from 100--200 m, and a drop off after 200 m. This would seem to indicate that the presence of the observers caused either movement away from the station (the crest around 100 m) or alternatively a suppression of vocalizations and/or movement directly around the station.

@@

One of the assumptions of distance sampling is that the objects being studied are distributed homogeneously with some uniform density, $D$, throughout the space \textcite{ramsey1981}.  (If the data is clustered we have other problems, not addressed here.) If the uniform distribution is the case, then it would make sense that it should not matter if the observations overlap, as long as the act of observing at one station does not affect the observations at the neighboring station. This echoes Buckland’s statement a few paragraphs previous.

@@

\subsection{Bias \& Consistency}
\textcite{buckland2001,ramsey1979} Buckland et al. (2001) and Ramsey \& Scott (1970) both claim that if our detectability estimate $g(z)=$1 out to $r$, then $\hat{D}$ is unbiased.

\textcite{barry2001} compared design-based and  model-based estimates of density and variance. (Their focus was line transects, but they state it is analogous to point transects and the conclusions should carry across.) They concluded that model-based estimates were unbiased if and only if the assumptions of objects being independent and uniformly located held. They noted that, in practice, this is probably not a feasible assumption. For design-based inference, they caution about both the density and variance estimates.

\textcite{roeder1987} use simulations to demonstrate that no method is completely applicable in 100\% of situations, and will go wrong at some point, depending on the specific conditions.

The overall conclusion is that, if all assumptions hold, density estimates using VCPs are unbiased and reasonably consistent. In practice, however, having the assumptions actually hold true is a precarious proposition. The appropriate estimation of $g(r)$ will vary depending on the specific situation, and it is up to the scientist and statistician to understand the data and choose the best approach.

\section{Simulation}
As was discussed in the previous section, the issue of independence of the point transects does not have a consensus in the literature. If independence is required for proper estimation, then the random-systematic placement (hereafter, ``transect'') layout may require some adjustment of the estimate to bring it in line with survey layouts where the observations areas do not overlap.

Through simulation, we can explore what happens under different VCP placement schemes, detection densities, and object movement amounts, and compare the resulting density estimates to their known values to evaluate the performance of the estimators.

\subsection{VCP Arrangement}
\label{sec:layout}
To examine the effect of VCP overlap, we will look at three options for VCP arrangement:
\begin{itemize}
	\item Structured: VCPs placed so that observation distances do not overlap
	\item Random: VCPs placed completely at random
	\item Transect: Two transects of 18 stations each, placed according to \textcite{micronesian} design: 150 m between stations, 2 km distance between parallel transects.
\end{itemize}
\begin{figure}
	\centering

	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{../images/layout_structured.pdf}
		\caption{Structured Layout}
		\label{fig:structured}
	\end{subfigure}
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{../images/layout_random.pdf}
		\caption{Totally Random Layout}
		\label{fig:random}
	\end{subfigure}

	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{../images/layout_rand-sys-4.pdf}
		\caption{A Transect Layout}
		\label{fig:transect1}
	\end{subfigure}
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{../images/layout_rand-sys-6.pdf}
		\caption{Another Transect Layout}
		\label{fig:transect2}
	\end{subfigure}
	\caption{VCP Layout options. On graph, 1 unit = 1 km. Circles represent 200 m mark.}
	\label{fig:layouts}
\end{figure}
Figures~\ref{fig:structured} and \ref{fig:random} illustrate the first two arrangements. Density in these images is based on the 20 birds per km$^2$ of the Palie region on Rota.

@@

\subsection{Detection Probability Options}
The first simulation compares the three layouts described in section~\ref{sec:layout} with two different options for $g(r)$, a half-normal detection probability function, and a detection probability estimated from the empirical observation distances.
\subsubsection{Half Normal}
For a maximum possible observation distance of 0.500 (500 m), the parameter for the half-normal detection function is $\theta = 8.77$, with a scale parameter $\delta = 0.179$.

@@

For a half-normal with parameter $\theta=8.7732$, $f(0)=5.5852$ so to scale the value to 1:
\begin{center}
$\delta_{HN} = 1/5.5852 = 0.1790$
\end{center}

@@

\section{Results}
Two simulations were run, one looking at the three layout options crossed with the two detection probability function options and the other examining the effect of the three layout options against the three movement options, holding the detection probability option standard.
\subsection{Half-normal vs. Empirical $g(r)$}
\begin{figure}

@@

The combination of compounded movement and the transect layout seems to have an effect on the density estimate. The other estimates for either movement type, are all in the same ballpark, around 11.8 with around a 59\% rate for capturing the true population value. For the Transect Layout and Compounded movement, this rises 0.50 to a mean population density estimate of 12.42, and a capture percent of 65\%. It is unclear if there is something about this combination that is reducing bias (possible, but feels unlikely), or if it is instead biasing the estimate upwards, which appears like bias reduction because the estimate is already low.

%\texttt{TODO: Try to xplain the .50 jump in the estimate for transect/compounded. No theory or story atm.}

@@

\section{The Problem with Circles}
To test the effectiveness of the movement algorithm, the detection distances from the data where the objects moved was compared to the detection distances from the same data, assuming the objects stayed still. The movement data, (left panel, Figure~\ref{fig:moveTest}) is squashed more towards 0.1 Km than the still data (right panel). But there is a problem. The still data does not show the nice shoulder at 0 that the literature instructs us to expect.

@@

The problem is because circles are different than rectangles. With a line transect, as we move away from the line an equal distance in either direction, the distance surveyed increases linearly. With a point transect, as we move away from the point, the distance surveyed increases exponentially. Figure~\ref{fig:perfectArea} shows the relationship for a rectangle and circle with the same area.

Figure~\ref{fig:perfectCount} shows the difference in the expected number of observed objects at distance $r$ given a line transect vs. a VCP. For the VCP, Area was calculated as $2\pi r$. For the line transect, Area was calculated as $(2r)L$, with $L=0.785$ chosen so that the area of the circle and rectangle were equal when $r=0.5$.

@@

The take-away point is that for a VCP survey, even if everything is perfect, the empirical detection distances will not show the nice shouldered layout of the ideal detection probability curve. \textcite{ramsey1981} speak about this, when comparing the ``CUM-D'' curve, or the plot of the cumulative number of observations at distance $r$ for both line and point transects. However, almost all the illustrations on how the collected data should look, shows data that echoes the detection probability curve: the half-bell-shape with a nice shoulder. Since this shape is impossible for VCP, explicit examples should be shown in the literature so researchers do not assume their is something wrong with their data.

Many researchers have worked on distance sampling, and this effect is taken into account in the calculation of population density estimates. When analyzing point transect data we take the derivative of $g(r)$, an adjustment which is not necessary in line transect surveys. It can be shown that this solution adjusts for the difference in how area is calculated, to result in correct density estimates.

@@

\begin{figure}
	\centering
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{../images/rect-circ-area.pdf}
		\caption{Area of rectangle vs circle at same distance from point/line}
		\label{fig:perfectArea}
	\end{subfigure}
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{../images/rect-circ-detection.pdf}
		\caption{Estimated detection numbers for rectangle (solid) vs circle (dashed)}
		\label{fig:perfectCount}
	\end{subfigure}
	\caption{Rectangles (Line Transect, solid) vs. Circles (Point Transect, dashed)}
	\label{fig:linevs}
\end{figure}

\section{Conclusions}
If all else is held equal, VCP layout does not seem to play a large role in population density estimates using the kernel method as described by \textcite{quang1993}. Violations of the expected detection probability curve, or by movement of the objects away from the observer play more of a role in biasing the resulting estimate.

All of the population density estimates were biased low using Quang's formula for $\hat{D}$. This bias was observed even when the detection function was a half-normal curve and analyzed with a normal kernel, and the VCPs were placed so that the observation area did not overlap. The bias was not severe in this case, it estimated a population density around 18 objects per km$^2$, when the true estimate was 20/km$^2$, but it is noticeable. Dr. Quang's paper does offer a bias-corrected estimate, which was not used in this simulation.

%\section{Bibliography}
